\documentclass[11pt]{article}

\usepackage[T1]{fontenc}
\usepackage{geometry}
\usepackage{amsmath, amssymb, amsthm}
\usepackage[scr]{rsfso}
\usepackage{bm}
\usepackage[%
    hidealllines=true,%
    innerbottommargin=15,%
    nobreak=true,%
]{mdframed}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{hyperref}

\geometry{a4paper, margin=1in, headheight=14pt}

\pagestyle{fancy}
\fancyhf{}
\renewcommand\headrulewidth{0.4pt}
\fancyhead[L]{\scshape MA3201: Topology}
\fancyhead[R]{\scshape \leftmark}
\rfoot{\footnotesize\it Updated on \today}
\cfoot{\thepage}

\newcommand{\C}{\mathbb{C}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}

\newcommand{\RP}{\R\mathrm{P}}

\newcommand{\norm}[1]{\Vert #1 \Vert}

\DeclareMathOperator{\diam}{diam}
\DeclareMathOperator{\supp}{supp}

\newmdtheoremenv[%
    backgroundcolor=blue!10!white,%
]{theorem}{Theorem}[section]
\newmdtheoremenv[%
    backgroundcolor=violet!10!white,%
]{corollary}{Corollary}[theorem]
\newmdtheoremenv[%
    backgroundcolor=teal!10!white,%
]{lemma}[theorem]{Lemma}

\theoremstyle{definition}
\newmdtheoremenv[%
    backgroundcolor=green!10!white,%
]{definition}{Definition}[section]
\newmdtheoremenv[%
    backgroundcolor=red!10!white,%
]{exercise}{Exercise}[section]

\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{example}{Example}
\newtheorem*{solution}{Solution}

\surroundwithmdframed[%
    linecolor=black!20!white,%
    hidealllines=false,%
    innertopmargin=5,%
    innerbottommargin=10,%
    skipabove=0,%
    skipbelow=0,%
]{example}

\numberwithin{equation}{section}

\title{
    \Large\textsc{MA3201} \\
    \Huge \textbf{Topology} \\
    \vspace{5pt}
    \Large{Spring 2022}
}
\author{
    \large Satvik Saha
    \\\textsc{\small 19MS154}
}
\date{\normalsize
    \textit{Indian Institute of Science Education and Research, Kolkata, \\
    Mohanpur, West Bengal, 741246, India.} \\
}

\begin{document}
    \maketitle

    \tableofcontents

    \section{Introduction}
    
    \subsection{Topological spaces}

    \begin{definition}
        A topology on some set $X$ is a family $\tau$ of subsets of $X$,
        satisfying the following.
        \begin{enumerate}
            \itemsep0em
            \item $\emptyset, X \in \tau$.
            \item All unions of elements from $\tau$ are in $\tau$.
            \item All finite intersections of elements from $\tau$ are in $\tau$.
        \end{enumerate}
        The sets from $\tau$ are declared to be open sets in the topological space
        $(X, \tau)$.
    \end{definition}
    \begin{example}
        Any set $X$ admits the indiscrete topology $\tau_{id} = \{\emptyset, X\}$, as
        well as the discrete topology $\tau_{d} = \mathcal{P}(X)$. Both of these are
        trivial examples.
    \end{example}
    \begin{example}
        Let $X$ be a set. The cofinite topology on $X$ is the collection of
        complements of finite sets, along with the empty set. Note that when $X$ is
        finite, this is simply the discrete topology.
    \end{example}

    \begin{definition}
        Let $\tau, \tau'$ be two topologies on the set $X$. We say that $\tau$ is
        finer than $\tau'$ if $\tau$ has more open sets than $\tau'$. In such a case,
        we also say that $\tau'$ is coarser than $\tau$.
    \end{definition}


    \subsection{Topological bases}

    \begin{definition}
        Let $(X, \tau)$ be a topological space. We say that $\beta \subseteq
        \tau$ is a base of the topology $\tau$ such that every open set $U \in \tau$
        is expressible as a union of elements from $\beta$. 
    \end{definition}

    \begin{definition}
        Let $X$ be a set, and let $\beta$ be a collection of subsets of $X$
        satisfying the following.
        \begin{enumerate}
            \itemsep0em
            \item For every $x \in X$, there exists $x \in B \in \beta$.
            \item For every $x \in X$ such that $x \in B_1 \cap B_2$, $B_1, B_2 \in
            \beta$, there exists $B \in \beta$ such that $x \in B \subseteq B_1 \cap
            B_2$.
        \end{enumerate}
        Then, $\beta$ generates a topology on $X$, namely the collection of all
        unions of elements of $\beta$.
    \end{definition}

    \begin{lemma}
        Let $\tau$ be a topology on $X$, and let $\beta \subseteq \tau$ be a
        collection of open sets. Then, $\beta$ is a basis of $\tau$, or generates
        $\tau$, if for every $x \in U \in \tau$, there exists $B \in \beta$ such that
        $x \in B \subseteq U$.
    \end{lemma}
    \begin{example}
        The collection of all open balls in $\R^n$ form a basis of the usual topology.
    \end{example}

    \begin{lemma}
        Let $X$ be equipped with the topologies $\tau$ and $\tau'$, and let $\beta$
        and $\beta'$ be the respective bases of these topologies. Then, $\tau$ is
        finer than $\tau'$ if and only if given $x \in B' \in \beta'$, there exists
        $x \in B \in \beta$ such that $B \subseteq B'$.
    \end{lemma}
    \begin{example}
        The collections of open balls in $\R^n$ generate the same topology as the
        collection of all open rectangles in $\R^n$.
    \end{example}
    \begin{example}
        Consider the topologies on $\R$ generated by the following bases.
        \begin{enumerate}
            \itemsep0em
            \item $\beta_1 = \{(a, b): a, b \in \R, a < b\}$.
            \item $\beta_2 = \{[a, b): a, b \in \R, a < b\}$.
            \item $\beta_3 = \{(a, b): a, b \in \R, a < b\} \cup \{(a, b) \setminus
            K\}$ where $K = \{1 / n: n \in \Z\}$.
        \end{enumerate}
        We call the topology generated by $\beta_2$ the lower limit topology, denoted
        $\R_\ell$. The topology generated by $\beta_3$ is denoted $\R_K$. Both of
        these are strictly finer than the standard topology.
    \end{example}

    \begin{definition}
        A sub-basis for some topology on $X$ is a collection $\rho$ of subsets of $X$
        whose union is the whole of $X$. The topology generated by $\rho$ is defined
        to be the topology generated by the collection of all finite intersections of
        elements of $\rho$.
    \end{definition}


    \subsection{Product topology}
    
    \begin{definition}
        Let $(X_1, \tau_1)$, $(X_2, \tau_2)$ be topological spaces. Then $\tau_1
        \times \tau_2$ generates the product topology on $X_1 \times X_2$.
    \end{definition}
    \begin{example}
        The product topology on $\R \times \R$, where $\R$ is equipped with the
        standard topology, coincides with the standard topology on $\R^2$.
    \end{example}

    \begin{lemma}
        If $\beta_1, \beta_2$ are bases of the topologies $\tau_1, \tau_2$, then
        $\beta_1 \times \beta_2$ and $\tau_1 \times \tau_2$ generate the same product
        topology.
    \end{lemma}
    \begin{proof}
        Given $(x_1, x_2) \in U$ where $U \subseteq X_1\times X_2$ is open in the
        product topology, recall that $U$ can be written as a union of the basic open
        sets $U_{1i} \times U_{2i}$, where $U_{1i} \in \tau_1$ and $U_{2i} \in
        \tau_2$. Suppose that $(x_1, x_2) \in U_1 \times U_2$. Thus, we can choose
        $B_1 \in \beta_1$, $B_2 \in \beta_2$ such that $x_1 \in B_1 \subseteq U_1$
        and $x_2 \in B_2 \subseteq U_2$. Thus, $(x_1, x_2) \in B_1 \times B_2
        \subseteq U_1 \times U_2 \subseteq U$.
    \end{proof}

    \begin{definition}
        The projection maps are defined as $\pi_i\colon X_1 \times \cdots X_k \to
        X_i$, $(x_1, \dots, x_k) \mapsto x_i$.
    \end{definition}

    \begin{lemma}
        The collection of elements of the form $\pi_1^{-1}(U_1)$ or
        $\pi_2^{-1}(U_2)$, where $U_1 \in \tau_1$ and $U_2 \in \tau_2$, forms a
        sub-basis of the product topology on $X_1 \times X_2$.
    \end{lemma}
    \begin{proof}
        Note that $\pi_1^{-1}(X_1) = X_1 \times X_2$. Now it is easy to see that
        finite intersections of elements of the form $U_1 \times X_2$ or $X_1 \times
        U_2$ where $U_1, U_2$ are open, are all of the form $U_1 \times U_2$ which is
        precisely a basis of the product topology.
    \end{proof}
    \begin{corollary}
        We can restrict ourselves to the sub-basis of elements of the form
        $\pi_1^{-1}(B_1)$ or $\pi_2^{-1}(B_2)$, where $B_1 \in \beta_1$, $B_2 \in
        \beta_2$ for some bases $\beta_1$, $\beta_2$ of $\tau_1, \tau_2$.
    \end{corollary}


    \subsection{Subspace topology}
    
    \begin{definition}
        Let $(X, \tau)$ be a topological space, and let $Y \subset X$. Then the
        collection $U \cap Y$ for all $U \in \tau$ comprises the subspace topology
        $\tau_Y$ on $Y$ induced by the topology $\tau$ on $X$.
    \end{definition}

    \begin{lemma}
        If $\beta$ is a basis for the topology on $X$, and $Y \subset X$, then the
        collection $B \cap Y$ for all $B \in \beta$ generates the subspace topology
        on $Y$.
    \end{lemma}

    \begin{lemma}
        An open set of $Y$ is open in $X$ if $Y$ is open in $X$.
    \end{lemma}
    \begin{proof}
        Let $U \subset Y$ be open in $Y$, then $U = V \cap Y$ for some open set $V$
        in $X$. If additionally $Y$ is open in $X$, this immediately shows that $U$
        is open in $X$.
    \end{proof}

    \begin{theorem}
        Let $(X, \tau_X)$, $(Y, \tau_Y)$ be topological spaces, and let $A \subseteq
        X$, $B \subseteq Y$. Then, there are two ways of assigning a natural topology
        on $A \times B$.
        \begin{enumerate}
            \itemsep0em
            \item Take the product topology on $X \times Y$, and consider the
            subspace topology induced by it on $A \times B$.
            \item Take the subspace topologies on $A$ induced by $\tau_X$, $B$
            induced by $\tau_Y$, and consider the product topology generated by them
            on $A \times B$.
        \end{enumerate}
        These two methods generate the same topology on $A \times B$.
    \end{theorem}
    \begin{proof}
        Open sets in 1 look like $(U \times V) \cap (A \times B)$, where $U \in
        \tau_X$, $V \in \tau_Y)$. Open sets in 2 look like $(U' \cap A) \times (V' \cap
        B)$, where $U' \in \tau_X$, $V' \in \tau_Y$, which can be rewritten as $(U'
        \times V') \cap (A \times B)$. It is easy to see that these describe
        precisely the same sets.
    \end{proof}


    \subsection{Order topology}

    \begin{definition}
        Let $X$ be a set with a simple order $<$. Then the collection of sets of the
        form $(a, b)$, $[a_0, b)$, $(a, b_0]$ where $a_0$ is the minimal element of
        $X$, $b_0$ is the maximal element of $X$, generate the order topology on
        $X$.
    \end{definition}
    \begin{example}
        The order topology on $\N$ is precisely the discrete topology.
    \end{example}

    \begin{definition}
        Let $X_1, X_2$ be simply ordered sets. The dictionary order on $X_1 \times
        X_2$ is defined as follows: $(x_1, x_2) < (y_1, y_2)$ if $x_1 < y_1$, or if
        $x_1 = y_1$ and $x_2 < y_2$.
    \end{definition}
    
    \begin{example}
        Consider $X = \{1, 2\} \times \N$, where both $\{1, 2\}$ and $\N$ are endowed
        with the discrete topology. Note that the product topology on $X$ is the
        discrete topology.

        Now consider the dictionary order on $X$. Here, $(1, 1)$ is the smallest
        element, so we can list the elements of $X$ in ascending order. Note that
        every $(1, m) < (2, n)$, for all $m, n \in \N$. Now, note that all singletons
        $\{(1, m)\}$ are open in the order topology on $X$. The same is true for the
        singletons $\{(1, n)\}$ for all $n > 1$. However, the singleton $\{(2, 1)\}$
        is \emph{not} open in the order topology.
    \end{example}

    \begin{example}
        Consider $\R$ with the usual topology, and $X = [0, 1) \cup \{2\}$. Then,
        $\{2\}$ is open in the subspace topology on $X$, but it is not open in the
        order topology on $X$.
    \end{example}

    \begin{lemma}
        The open rays of the form $(a, +\infty)$ and $(-\infty, a)$ in $X$ form a
        sub-basis of the order topology on $X$.
    \end{lemma}
    \begin{proof}
        Note that $(a, b) = (-\infty, b) \cap (a, +\infty)$, $[a_0, b) = (-\infty,
        b)$, and $(a, b_0] = (a, +\infty)$.
    \end{proof}

    \begin{definition}
        Let $X$ be a simply ordered set, and $Y \subseteq X$. Then, we say that $Y$
        is convex in $X$ if given $a, b \in Y$ such that $a < b$, the interval $(a,
        b) = \{x \in X: a < x < b\} \subseteq Y$.
    \end{definition}

    \begin{theorem}
        Let $Y$ be convex in $X$. Then, the subspace topology and the order topology
        on $Y$ induced from the order topology on $X$ coincide.
    \end{theorem}


    \subsection{Closed sets}
    
    \begin{definition}
        Let $(X, \tau)$ be a topological space. A set $F \subseteq X$ is said to be
        closed in $X$ if $F^c = X\setminus F \in \tau$.
    \end{definition}
    \begin{example}
        The sets $\emptyset, X$ are closed in every topological space $(X, \tau)$.
    \end{example}
    \begin{example}
        In a set equipped with the discrete topology, every set is both open and
        closed.
    \end{example}

    \begin{lemma}
        Arbitrary intersections, and finite unions of closed sets are closed.
    \end{lemma}

    \begin{theorem}
        Let $(X, \tau)$ be a topological space, and let $Y \subset X$ be equipped
        with the subspace topology. Then, a set $F \subseteq Y$ is closed in $Y$ if
        and only if $F = Y \cap G$, where $G$ is closed in $X$.
    \end{theorem}
    \begin{proof}
        Let $F \subset Y$. Now, $F$ is closed in $Y$, $Y\setminus F = Y \cap F^c$ is
        open in $Y$, $Y \cap F^c = Y \cap U$ where $U$ is open in $X$, $F =
        Y \cap (Y \cap F^c)^c = Y \cap (Y \cap U)^c = Y \cap U^c$ where $U^c$ is
        closed. The steps are reversible.
    \end{proof}

    \begin{lemma}
        A closed set of $Y$ is closed in $X$ if $Y$ is closed in $X$.
    \end{lemma}

    \subsection{Interiors and closures}
    
    \begin{definition}
        Let $A \subseteq X$ where $(X, \tau)$ is a topological space. \begin{enumerate}
            \itemsep0em
            \item The interior of $A$ is defined as the union of all open sets
            contained in $A$. This is denoted by $A^\circ$.
            \item The closure of $A$ is defined as the intersection of all closed
            sets containing $A$. This is denoted by $\overline{A}$.
        \end{enumerate}
        \begin{remark}
            The interior of a set is open, and the closure of a set is closed.
        \end{remark}
    \end{definition}
    
    \begin{lemma}
        Let $Y \subset X$ be topological spaces, and let $A \subseteq Y$. Also let
        $\overline{A}_X$, $\overline{A}_Y$ denote the closures of $A$ in $X$, $Y$
        respectively. Then, $\overline{A}_Y = \overline{A}_X \cap Y$.
    \end{lemma}

    \begin{theorem}
        Let $A \subset X$. Then, \begin{enumerate}
            \itemsep0em
            \item $x \in \overline{A}$ if and only if every open set containing $x$
            has non-empty intersection with $A$.
            \item $x \in \overline{A}$ if and only if every basic open set containing
            $x$ has non-empty intersection with $A$, given that the topology on $X$
            is generated by those basic open sets.
        \end{enumerate}
    \end{theorem}


    \begin{definition}
        Let $A \subseteq X$ where $(X, \tau)$ is a topological space. We say that $x
        \in X$ is a limit point of $X$ if for every open set $U$ containing $x$, the
        deleted neighbourhood $U \setminus\{x\}$ has non-empty intersection with $A$.
        The set of limit points of $A$ is denoted by $A'$.
    \end{definition}

    \begin{example}
        Let $X$ be a set endowed with the discrete topology. Then, given any set $A
        \subseteq X$, we have $A' = \emptyset$.
    \end{example}

    \begin{lemma}
        A closed set contains all its limit points.
    \end{lemma}
    \begin{proof}
        Let $F \subseteq X$ be closed in $X$, and let $x \in F'$. Then given any open
        set containing $x$, we have $U \cap F \supseteq (U\setminus\{x\}) \cap F \neq
        \emptyset$, hence $x \in \overline{F} = F$.
    \end{proof}

    \begin{lemma}
        Let $A \subseteq X$ where $(X, \tau)$ is a topological space. Then,
        $\overline{A} = A \cup A'$.
    \end{lemma}
    \begin{proof}
        It is clear that $\overline{A} \supseteq A \cup A'$. Now pick $x \in
        \overline{A}$. If $x \notin A$, then we know that given any open
        neighbourhood $U$ of $x$, we have non-empty $U \cap A$. Furthermore, this
        intersection can never contain $x$, hence $x \in A'$. This proves that
        $\overline{A} \subseteq A \cup A'$.
    \end{proof}


    \subsection{Convergence of sequences}

    \begin{definition}
        Let $\{x_n\}_{n = 1}^\infty$ be a sequence of points from $(X, \tau)$, and
        let $x \in X$. We say that this sequence converges to $x$, denoted $x_n \to
        x$, if every open neighbourhood of $x$ contains the tail of this sequence. In
        other words, given $U \in \tau$ such that $x \in U$, there must exist $N \in
        \N$ such that $x_n \in U$ for all $n \geq N$.
    \end{definition}

    \begin{example}
        Let $X = \{a, b, c\}$, and $\tau = \{\emptyset, \{b\}, \{a, b\}, \{b, c\},
        X\}$. Then, the constant sequence of $b$'s converges to all three points $a,
        b, c$.
    \end{example}
    \begin{example}
        Let $X = \R$, and $\tau$ be the collection of all intervals $(-a, a)$
        together with $\emptyset, \R$. Then, the constant sequence of $0$'s converges
        to every point in $\R$.
    \end{example}

    \begin{definition}
        Let $(X, \tau)$ be a topological space. We say that this topological space is
        Hausdorff if given any two distinct points $x, y \in X$, there exist open
        sets $U, V \in \tau$ such that $x \in U$, $y \in V$, and $U \cap V =
        \emptyset$.
    \end{definition}

    \begin{example}
        The real numbers under the standard topology is Hausdorff.
    \end{example}

    \begin{theorem}
        Let $(X, \tau)$ be a Hausdorff topological space, and let $\{x_n\}_{n =
        1}^\infty$ be a sequence of points in $X$. Then, this sequence can converge
        to at most one point in $X$.
    \end{theorem}
    \begin{proof}
        Suppose that $\{x_n\}_{n = 1}^\infty$ converges to distinct points $x, y \in
        X$. Then there exist disjoint open neighbourhoods $U, V$ such that $x \in U$,
        $y \in V$. Convergence means that both $U$ and $V$ contain a tail of the
        sequence, which is a contradiction.
    \end{proof}

    \begin{lemma}
        The singleton sets in a Hausdorff space are closed.
    \end{lemma}
    \begin{proof}
        Let $x \in X$ where $(X, \tau)$ is Hausdorff. Pick $y \neq x$, whence there
        exist $U_y, V_y \in \tau$, such that $x \in U_y$, $y \in V_y$, and $U_y \cap
        V_y = \emptyset$. In particular, $\{x\} \cap V_y = \emptyset$. We now have \[
            X\setminus \{x\} = \bigcup_{y \neq x} V_y,
        \] which is open.
    \end{proof}

    \begin{theorem}
        The topology induced by a metric is Hausdorff.
    \end{theorem}
    \begin{proof}
        Given a metric space $X$ and distinct points $x, y \in X$, we set $r = |x -
        y|$, $U = B(x, r/3)$, $V = B(y, r/3)$.
    \end{proof}
    

    \section{Continuous maps}

    \begin{definition}
        Let $f\colon X \to Y$ be a function between the topological spaces $(X,
        \tau_X)$ and $(Y, \tau_Y)$. We say that $f$ is continuous if for every $U
        \in \tau_Y$, we have $f^{-1}(U) \in \tau_X$. In other words, the pre-image of
        every open set in $Y$ must be open in $X$.
    \end{definition}

    \begin{lemma}
        A function $f\colon X \to Y$ is continuous if and only if given a base
        $\beta$ of $Y$, we have $f^{-1}(U) \in \tau_X$ for every $U \in \beta$.
    \end{lemma}
    \begin{example}
        The identity function $\operatorname{id}\colon \R_\ell \to \R$ is continuous, while the
        identity function $\operatorname{id}\colon \R \to \R_\ell$ is not. This is
        because the topology on $\R_\ell$ is strictly finer than that on $\R$.
    \end{example}

    \begin{lemma}
        A function $f\colon X \to Y$ is continuous if and only if for every closed
        set $F \subseteq Y$, we have $f^{-1}(F)$ closed in $X$.
    \end{lemma}

    \begin{lemma}
        A function $f\colon X \to Y$ is continuous if and only if given any $x \in X$
        and an open set $V \subseteq Y$ such that $f(x) \in V$, there exists an open
        set $U \subseteq X$ such that $x \in U$, $f(U) \subseteq V$.
    \end{lemma}

    \begin{theorem}
        The composition of continuous functions is continuous.
    \end{theorem}

    
    \subsection{Restricting and enlarging the domain}

    \begin{lemma}
        Let $f\colon X \to Y$ be continuous, and let $A \subset X$. Then the
        restriction of $f$ to $A$ is continuous.
    \end{lemma}

    \begin{theorem}
        Let $f\colon X \to Y$, and let $X$ be the union of the collection of open
        sets $\{A_\lambda\}_{\lambda \in \Lambda}$. If the restrictions of $f$ to
        each $A_\lambda$ are continuous, then $f$ is continuous.
    \end{theorem}
    \begin{proof}
        Pick $x \in X$, hence $x \in A_\lambda$ for some $\lambda \in \Lambda$. Now
        if $f(x) \in V \subset Y$, where $V$ is open in $Y$, then the continuity of
        the restriction of $f$ to $A_\lambda$ gives us an open set $U \subseteq
        A_\lambda$ such that $f(U) \subseteq V$. Finally since $A_\lambda$ is open in
        $X$, so is $U$.
    \end{proof}
    
    \begin{definition}
        Let $X$ be the union of the collection of open sets $\{A_\lambda\}_{\lambda
        \in \Lambda}$. We say that this collection is a locally finite cover of $X$
        if given $x \in X$, there exists a neighbourhood $U$ of $x$ such that $U \cap
        A_\lambda$ is non-empty for only finitely many $\lambda \in \Lambda$.
    \end{definition}
    
    \begin{theorem}
        Let $f\colon X \to Y$, and let $\{F_\lambda\}_{\lambda \in \Lambda}$ be a
        locally finite collection of closed sets covering $X$. If the restrictions of
        $f$ to each $F_\lambda$ are continuous, then $f$ is continuous.
    \end{theorem}

    \begin{corollary}[Pasting lemma]
        Let $X = A\cup B$, with $A, B$ closed in $X$. Let $f\colon A \to Y$, $g\colon
        B \to Y$ be continuous, with $f(x) = g(x)$ on $A \cap B$. Then the function
        $h\colon X \to Y$, defined by $x\mapsto f(x)$ on $A$ and $x\mapsto g(x)$ on
        $B$, is continuous.
    \end{corollary}

    \begin{definition}
        A path is a continuous function $\gamma\colon [0, 1] \to X$.
    \end{definition}

    \begin{lemma}
        Two paths $\gamma_1, \gamma_2$ can be concatenated when $\gamma_1(1) =
        \gamma_2(0)$.
    \end{lemma}


    \subsection{Homeomorphisms}

    \begin{definition}
        Let $f\colon X \to Y$ be a function between the topological spaces $(X,
        \tau_X)$ and $(Y, \tau_Y)$. We say that $f$ is a homeomorphism if $f$ is
        continuous, $f$ is bijective, and $f^{-1}$ is continuous. We also say that
        $X$ and $Y$ are homeomorphic when such a homeomorphism between them exists.
    \end{definition}
    \begin{example}
        The interval $(-1, 1)$ is homeomorphic to $\R$; for instance, the map $x
        \mapsto \tan(\pi x/ 2)$ on $(-1, 1)$ is a homeomorphism. A simpler
        construction is the map $x \mapsto x / (1 - x^2)$.
    \end{example}

    
    \subsection{Projection maps}

    \begin{theorem}
        The projection maps $\pi_i\colon X_1\times \dots \times X_k \to X_i$ are
        continuous, when the domain is equipped with the product topology.
        Furthermore, the product topology is the coarsest topology on the domain for
        which the projection maps are continuous.
    \end{theorem}

    \begin{lemma}
        Let $f\colon A \to X_1 \times \dots \times X_k$, where the co-domain is
        equipped with the product topology. Then, $f$ is continuous if and only if
        the component functions $f_i = \pi_i\circ f$ are continuous.
    \end{lemma}
    \begin{proof}
        Note that if $f$ is continuous, the compositions $\pi_i\circ f$ are
        immediately continuous. Conversely suppose that each $f_i$ is continuous, and
        write \[
            f(t) = (f_1(t), \dots, f_k(t)).
        \] The sets $U_1 \times \dots \times U_k$, where each $U_i \subseteq X_i$ is
        open, form a basis of the co-domain. Furthermore, their pre-images under $f$
        are $f_1^{-1}(U_1) \cap \dots \cap f_k^{-1}(U_k)$, which are open in $A$.
        This shows that $f$ is continuous.
    \end{proof}


    \begin{definition}
        Let $J$ be an arbitrary index set. A $J$-tuple of elements in a set $X$ is a
        function $x\colon J \to X$, formally denoted $(x_\alpha)_{\alpha \in J}$.
        If $\{X_\alpha\}_{\alpha \in J}$ is a family of sets, their Cartesian product
        is defined as \[
            \prod_{\alpha \in J} X_\alpha = \{x\colon J \to \bigcup_{\alpha \in J}
            X_\alpha\colon x_\alpha \in X_\alpha\}.
        \] 
        \begin{remark}
            The fact that we can choose an element from each set in an uncountable
            collection relies on the Axiom of Choice.
        \end{remark}
    \end{definition}

    \begin{definition}
        Let $\{X_\alpha\}_{\alpha \in J}$ be a collection of topological spaces. The
        topology generated by $\prod_{\alpha \in J} U_\alpha$, where each $U_\alpha
        \subseteq X_\alpha$ is open, is called the box topology on $\prod_{\alpha \in
        J} X_\alpha$.
    \end{definition}
    
    \begin{definition}
        Let $\{X_\alpha\}_{\alpha \in J}$ be a collection of topological spaces. The
        topology generated by the sub-basis $\pi_\alpha^{-1}(U_\alpha)$, where each
        $U_\alpha \subseteq X_\alpha$ is open, is called the product
        topology on $\prod_{\alpha \in J} X_\alpha$.

        \begin{remark}
            The basic open sets are of the form $\pi_{\alpha \in J} U_\alpha$, where
            all but finitely many $U_\alpha = X_\alpha$. Thus, this is a coarser
            topology than the box topology.
        \end{remark}
    \end{definition}

    \begin{lemma}
        Let $\prod_{\alpha \in J}X_\alpha$ be equipped with the box or product
        topology. Then, $\overline{\prod A_\alpha} = \prod \overline{A}_\alpha$,
        where each $A_\alpha \in X_\alpha$.
    \end{lemma}
    
    \begin{lemma}
        Let $f\colon A \to \prod_{\alpha \in J} X_\alpha$, where the co-domain is
        equipped with the product topology. Then, $f$ is continuous if and only if
        the component functions $f_\alpha = \pi_\alpha\circ f$ are continuous.

        \begin{remark}
            This fails when $\prod_{\alpha \in J}$ is equipped with the box topology.
            Consider $f\colon \R \to \prod_{n = 1}^\infty \R$, $x \mapsto (x, x,
            \dots)$. Then, the product $\prod_{n = 1}^\infty (- 1 / n, 1 / n)$ is
            open in the box topology, but its pre-image under $f$ is $\bigcap_{n =
            1}^\infty (-1 / n, 1 / n) = \{0\}$, which is not open in $\R$.
        \end{remark}
    \end{lemma}
    

    \section{Metric spaces}
    
    \begin{definition}
        A metric space $(X, d)$ is a set equipped with a metric $d\colon X \times X
        \to \R$, such that \begin{enumerate}
            \itemsep0em
            \item $d(x, y) = 0$ if and only if $x = y$.
            \item $d(x, y) = d(y, x)$.
            \item $d(x, z) \leq d(x, y) + d(y, z)$.
        \end{enumerate}
    \end{definition}

    \begin{definition}
        An open ball in a metric spaces is the set of points \[
            B(x, r) = \{y \in X : d(x, y) < r\}.
        \] 
    \end{definition}

    \begin{lemma}
        The collection of open balls in a metric space generates its standard
        topology.
    \end{lemma}

    \begin{example}
        Consider a set $X$, equipped with the metric \[
            d\colon X \times X \to \R, \qquad (x, y) \mapsto \begin{cases}
                0, &\text{ if } x = y, \\
                1, &\text{ if } x \neq y.
            \end{cases}
        \] Then, this metric induces the discrete topology on $X$.
    \end{example}


    \subsection{Metrizable spaces}
    
    \begin{definition}
        A topological space $(X, \tau)$ is called metrizable if there exists a metric
        $d\colon X \times X \to \R$ which induces the topology $\tau$ on $X$.
    \end{definition}


    \begin{definition}
        Let $A \subseteq X$. The diameter of $A$ is defined to be \[
            \diam(A) = \sup\{d(x, y): x, y \in A\}.
        \] If $\diam(A)$ is finite, we say that $A$ is bounded.
    \end{definition}
    \begin{example}
        The metric \[
            (x, y) \mapsto \frac{|x - y|}{1 + |x - y|}
        \] generates the standard topology on $\R$. Note that $\R$ is unbounded in
        the standard metric, but bounded in this one.
    \end{example}

    \begin{definition}
        Let $(X, d)$ be a metric space. Then the standard bounded metric
        corresponding to $d$ is defined as \[
            \bar{d}\colon X \times X \to \R, \qquad (x, y) \mapsto \min\{d(x, y), 1\}.
        \] 
    \end{definition}

    \begin{lemma}
        Both $d$ and $\bar{d}$ generate the same topology.
    \end{lemma}


    \begin{theorem}
        The product topology on $\R^\omega = \R \times \R \times \dots$ is
        metrizable, using the metric \[
            D(x, y) = \sup_n\left\{\frac{1}{n}\bar{d}(x, y)\right\}.
        \] 
    \end{theorem}

    \begin{lemma}[Sequence lemma]
        Let $A \subseteq X$, let $x \in X$, and let the sequence $\{x_n\}_{n \in
        \N}$, $x_n \in A$ converge with $x_n \to x$. Then, $x \in \overline{A}$.
        \begin{remark}
            The converse holds if $X$ is metrizable.
        \end{remark}
    \end{lemma}

    \begin{example}
        Consider $X = \R^\omega$ equipped with the box topology. Choose $A = \{(x_1,
        x_2, \dots): x_i > 0\}$. Then, $0 = (0, 0, \dots) \in \overline{A}$; this is
        clear from the fact that any open set around $0$ contains the basic open set
        $\prod_i (a_i, b_i)$ with $a_i < 0 < b_i$. However, there is no sequence
        $\{x_n\}_{n \in \N}$, $x_n \in A$, such that $x_n \to 0$. Note that if this
        were the case, then each $x_n = (x_{n1}, x_{n2}, \dots)$. Now, $B = \prod_i
        (-x_{ii}, x_{ii})$ contains none of the points $x_n$, since the $n$th
        coordinate of $B$ eliminates the point $n$. \\
    \end{example}
    \begin{corollary}
        $\R^\omega$ equipped with the box topology is not metrizable.
    \end{corollary}


    \section{Compactness}
    
    \begin{definition}
        Let $X$ be a topological space. We say that $X$ is compact if every open
        cover of $X$ has a finite subcover.
    \end{definition}

    \begin{lemma}
        Let $Y \subseteq X$. Then, $Y$ is compact if and only if every open cover of
        $Y$ by open sets in $X$ has a finite subcover.
    \end{lemma}


    \subsection{Compact subspaces}
    
    \begin{lemma}
        All compact sets in a metric space are bounded.
    \end{lemma}
    \begin{proof}
        Let $K \subseteq X$ be compact. Then, $K$ admits an open cover of open balls
        $B(0, n)$ from which we can extract a finite subcover; however, this can be
        reduced to just one open ball $B(0, N)$ for some $N$. Thus $K \subset B(0,
        N)$ is bounded.
    \end{proof}

    \begin{lemma}
        A closed subset of a compact space is compact.
    \end{lemma}
    \begin{proof}
        Let $K$ be compact, and $F \subseteq K$ be closed. Consider an open cover
        $\{U_\alpha\}_{\alpha \in J}$ of $F$. By adding $K\setminus F$ to this
        collection, we have an open cover of $K$, from which we can extract a finite
        subcover $U_{i_1}, U_{i_2}, \dots, U_{i_k}, K\setminus F$. By discarding the
        latter, we have found a finite subcover of $F$.
    \end{proof}

    \begin{lemma}
        In a Hausdorff space, every compact set is closed.
    \end{lemma}
    \begin{proof}
        Let $X$ be Hausdorff, and $K \subseteq X$ be compact. Fix $x_0 \in X\setminus
        K$, and note that given any $y \in K$, there exist open neighbourhoods $U_y,
        V_y$ such that $x_0 \in U_y$, $y \in V_y$, $U_y \cap V_y = \emptyset$. Thus,
        the collection of all such $\{V_y\}_{y \in K}$ is an open cover of $K$, from
        which we can extract a finite subcover $V_{y_1}, \dots, V_{y_k}$.
        Corresponding to this, $x_0 \in U_{y_1} \cap \dots \cap U_{y_k} \subseteq X
        \setminus K$. Thus, $x_0$ lies in the interior of $X\setminus K$. This shows
        that $X \setminus K$ is open, hence $K$ is closed.
    \end{proof}

    \begin{theorem}
        The image of a compact space under a continuous map is compact.
    \end{theorem}

    \begin{lemma}
        Let $f\colon X \to Y$ be a continuous bijection. If $X$ is compact and $Y$ is
        Hausdorff, then $f$ is a homeomorphism.
    \end{lemma}
    \begin{proof}
        We need only show that $f$ is a closed map; now every closed set $F \subseteq
        X$ is compact because $X$ is compact, hence $f(K) \subseteq Y$ is compact.
        Since $Y$ is Hausdorff, the compact set $f(K)$ is closed.
    \end{proof}

    \begin{lemma}
        A topological space is compact if and only if for any collection of closed
        sets $\mathscr{C}$ with the finite intersection property, we have $\bigcap_{C
        \in \mathscr{C}} C \neq \emptyset$.
    \end{lemma}


    \subsection{Products of compact spaces}

    \begin{lemma}[Tube lemma]
        Let $X, Y$ be topological spaces, and let $Y$ be compact. Let $x_0 \in X$,
        and let $\{x_0\} \times Y \subset N \subseteq X \times Y$ where $N$ is open.
        Then, there exists an open set $W \subseteq X$ such that $\{x_0\} \times Y
        \subseteq W \times Y \subseteq N$.
    \end{lemma}
    \begin{proof}
        Note that $\{x_0\}\times Y$ is compact, being homeomorphic to $Y$. Thus, it
        can be covered with basic open sets $U_1\times V_1, \dots, U_k\times V_k$
        such that each $U_i \times V_i \subset N$. Simply set $W = U_1 \cap \dots
        \cap U_k$.
    \end{proof}

    \begin{theorem}
        Let $X, Y$ be compact topological spaces. Then, $X \times Y$ is compact.
    \end{theorem}
    \begin{proof}
        Let $\{U_\alpha\}_{\alpha \in J}$ be an open cover of $X \times Y$. Pick $x
        \in X$, whence $\{x\} \times Y$ is compact and admits a finite subcover
        $U_{xi_1}, \dots, U_{xi_k}$. Denote their union by $U_x$; the tube lemma
        guarantees an open set $W_x \subseteq X$ such that $\{x\} \times Y \subseteq
        W\times Y \subseteq U_x$. Now, the collection $\{W_x\}_{x \in X}$ is an open
        cover of $X$, hence admits a finite subcover $W_{x_1}, \dots, W_{x_n}$. This
        also means that $W_{x_1}\times Y, \dots, W_{x_n} \times Y$ is a finite cover
        of $Y$. However, each $W_{x_i} \times Y \subseteq U_{x_i}$ can be covered by
        finitely many $U_\alpha$, which means that we have a finite subcover of $X
        \times Y$.
    \end{proof}


    \subsection{Euclidean spaces}

    \begin{lemma}
        Let $X$ be a simply ordered set with the least upper bound property. Then,
        the intervals $[a, b]$ are compact.
    \end{lemma}

    \begin{theorem}[Heine-Borel]
        Compact sets of $\R^n$ are precisely those which are closed and bounded.        
    \end{theorem}


    \subsection{Limit point compactness}

    \begin{definition}
        Let $X$ be a topological space. We say that $X$ is limit point compact if
        every infinite subset of $X$ has a limit point.
    \end{definition}

    \begin{lemma}
        A compact space is limit point compact.
    \end{lemma}
    \begin{proof}
        Let $X$ be compact, and let $A \subseteq X$ have no limit points. Then, $A =
        A \cup A' = \overline{A}$ is closed in $X$, hence compact. Now given any $a
        \in A$, we know that $a$ is not a limit point of $A$, hence we can choose an
        open neighbourhood $U_a$ such that $U_a \cap A = \{a\}$. The collection
        $\{U_a\}_{a \in A}$ is now an open cover of $A$, and hence admits a finite
        subcover $U_{a_1}, \dots, U_{a_k}$. Let $U$ denote their union, whence $A = A
        \cap U = \{a_1, \dots, a_k\}$ is finite.
    \end{proof}
    \begin{example}
        Let $X = \N \times \{0, 1\}$, where $\N$ has the discrete topology, and $\{0,
        1\}$ has the indiscrete topology. Then, every subset of $X$ has a limit
        point; indeed, given any $\{(n, b)\}$, we have a limit point $(n, 1 - b)$.
        However, $X$ is clearly not compact, since the open cover of sets
        $\{n\}\times \{0, 1\}$ does not admit any finite subcover.
    \end{example}

    \begin{theorem}
        Let $X$ be a metrizable space. Then, $X$ is limit point compact if and only
        if it is compact.
    \end{theorem}


    \section{Connectedness}
    
    \begin{definition}
        Let $X$ be a topological space, and let $U, V \subseteq X$ be open,
        non-empty, disjoint, with $U \cup V = X$. We say that $U, V$ form a
        separation of $X$.
    \end{definition}

    \begin{definition}
        A topological space $X$ is said to be connected if it admits no separation.
    \end{definition}

    \begin{lemma}
        A topological space $X$ is connected if and only if the only subsets that are
        both open and closed in it are $\emptyset, X$.
    \end{lemma}

    \begin{lemma}
        Let $X$ be a topological space, and let $Y \subseteq X$ be a subspace. Then,
        a separation of $Y$ is a pair of open sets $A, B \subseteq X$ such that
        $\overline{A} \cap B = \emptyset$, $A \cap \overline{B} = \emptyset$.
    \end{lemma}

    \begin{lemma}
        Let $C, D$ form a separation of $X$, and let $Y \subseteq X$ be a connected
        subspace. Then, either $Y \subseteq C$, $Y \subseteq D$.
    \end{lemma}

    \begin{lemma}
        The union of a collection of connected spaces with a common point is
        connected.
    \end{lemma}
    \begin{proof}
        Let $\{X_\alpha\}_{\alpha \in J}$ be a collection of connected spaces, with
        the common point $x_0$, and let $X$ be their union. Suppose that $U, V$ is a
        separation of $X$; then each of the connected $X_\alpha$ must be contained in
        one of $U, V$. However, since all $X_\alpha$ share the common point $x_0$,
        they must all lie in the same half, say $U$, forcing $V = \emptyset$, a
        contradiction.
    \end{proof}
    

    \begin{lemma}
        Let $A \subseteq X$ be connected, and let $A \subseteq B \subseteq
        \overline{A}$. Then, $B$ is connected.
    \end{lemma}


    \begin{theorem}
        The image of a connected space under a continuous maps is connected.
    \end{theorem}

    \begin{theorem}
        A finite Cartesian product of connected spaces is connected.
    \end{theorem}
    \begin{proof}
        Let $X, Y$ be connected spaces. Fix $(a, b) \in X \times Y$. Now, $X\times
        \{b\}$ is connected, being homeomorphic to $X$. Furthermore, each $\{x\}
        \times Y$ is connected, for each $x \in Y$. Now, the set $T_x = \{x\} \times
        Y \cup X\times \{b\}$ is connected, being the union of connected spaces with
        the common point $(x, b)$. Finally, the union of all such $T_x$ is connected,
        being the union of connected spaces with the common point $(a, b)$. This
        union is just $X\times Y$, which is thus connected.
    \end{proof}
    \begin{example}
        The countable product $\R^\omega$ with the box topology is disconnected.
        Consider \[
            A = \text{set of all bounded sequences}, \qquad
            B = \text{set of all unbounded sequences}.
        \] Now, $A \cap B = \emptyset$, $A\cup B = \R^\omega$, $A, B \neq \emptyset$.
        It can also be shown that $A, B$ are open.
    \end{example}
    \begin{example}
        The countable product $\R^\omega$ with the product topology is connected. To
        show this, define \[
            \tilde{R}^n = \{(x_1, \dots, x_n, 0, 0, \dots): x_i \in \R\}.
        \] Then, set $X = \bigcup_{n = 1}^\infty \tilde{R}^n$, and note that since
        each $\tilde{R}^n \cong \R^n$ is connected with all of them sharing the
        common point $(0, 0, \dots)$, $X$ must be connected. We now show that
        $\overline{X} = \R^\omega$. Indeed, given $x = (x_1, x_2, \dots) \in \R^\omega$,
        an open neighbourhood of $x$ looks like $U = U_1 \times U_2 \times \dots$,
        where all but finitely many $I_i = \R$. In other words, there exists
        sufficiently large $N in \N$ such that for all $n \geq N$, $U_n = \R$. Thus,
        the point $(x_1, x_2, \dots, x_n, 0, 0, \dots) \in U \cap \tilde{R}^N$.
    \end{example}

    \begin{lemma}
        The closed intervals $[a, b] \subset \R$ are connected.
    \end{lemma}

    \subsection{Path connectedness}

    \begin{definition}
        A topological space $X$ is said to be path connected if there exists a path
        joining any two points in $X$. In other words, given $a, b \in X$, there
        always exists a continuous map $\gamma\colon [0, 1] \to X$ such that
        $\gamma(0) = a$, $\gamma(1) = b$.
    \end{definition}

    \begin{lemma}
        All path connected spaces are connected.
    \end{lemma}
    \begin{proof}
        Note that if $X = U \cup V$ is a separation of the path connected space $X$,
        then $[0, 1] = \gamma^{-1}(U) \cup \gamma^{-1}(V)$ is a separation of the
        connected interval $[0, 1]$, a contradiction.
    \end{proof}

    \begin{lemma}
        The image of a path connected space under a continuous map is path connected.
    \end{lemma}
    \begin{example}
        The unit sphere $S^{n - 1}$ is path connected. Note that the map \[
            f\colon \R^n\setminus\{0\} \to S^{n - 1}, \qquad x \mapsto x / \norm{x}
        \] is continuous and surjective. Thus, it maps the path connected set
        $\R^n\setminus\{0\}$ to $S^{n - 1}$, which must be path connected.
    \end{example}
    \begin{example}
        The set $\overline{S}$, called the topologist's sine curve, is connected but
        not path connected. \[
            S = \{(x, \sin(1 / x)): 0 < x \leq 1\}.
        \] Note that $S$ is the continuous image of the connected interval $(0, 1]$,
        hence connected. This further shows that $\overline{S}$ is connected. Now, \[
            \overline{S} = S \cup \{(0, y): -1 \leq y \leq 1\}.
        \] However, $\overline{S}$ is not path connected, since there exists no path
        joining $(0, 0)$ and $(1 / \pi, 0)$. Indeed, given any path $\gamma\colon [0,
        1] \to \overline{S}$ starting at $(0, 0)$, it cannot escape $\{0\}\times [-1,
        1]$. To see this, write $\gamma = (\gamma_1, \gamma_2)$, $\gamma_2(0) = 0$.
        By continuity of $\gamma_2$, we can choose $\delta > 0$ such that
        $|\gamma_2(t)| < 1 / 2$ for all $0 \leq t \leq \delta$. Suppose that
        $\gamma_1(t^*) = \tau > 0$ for some $0 \leq t \leq \delta$. By the
        intermediate value theorem, $\gamma_1$ takes all the values between $0$ and
        $\tau$ in the interval $[0, t^*]$. Choose $N$ such that $2 / \pi(2N + 1) <
        \tau$. Again, there must exist some $0 < t_0 < t^*$ such that $\gamma_1(t_0)
        = 2 / \pi(2n + 1)$. Now, $\gamma_2(t_0) = \sin(1 / \gamma_1(t_0)) = 1 > 1 /
        2$, a contradiction. This means that $\gamma_1(t) = 0$ for all $t \in [0,
        \delta]$.
    \end{example}


    \section{Quotient topology}
    
    \begin{definition}
        Let $X$ be a topological space, and let $\sim$ be an equivalence relation on
        $X$. Then $X/\!\sim$ denotes the set of all equivalence classes with respect
        to $\sim$. Its elements are of the form $[x] = \{y \in X : x \sim y\}$, for
        $x \in X$. Define the map \[
            \pi\colon X \to X/\!\sim, \qquad x \mapsto [x].
        \] The quotient topology on $X/\sim$ is the finest topology such that $\pi$
        is continuous. In other words, $U \subseteq X/\!\sim$ is open if
        $\pi^{-1}(U)$ is open in $X$.
    \end{definition}

    \begin{lemma}
        Let $f\colon X \to Y$ be a continuous surjection, with $X$ compact and $Y$
        Hausdorff. Define an equivalence relation $\sim$ on $X$ such that $x\sim
        x'\Leftrightarrow f(x) = f(x')$. Then, $g\colon X/\!\sim \;\to Y$, $[x]
        \mapsto f(x)$ is a homeomorphism.
    \end{lemma}
    \begin{example}
        Consider the interval $[0, 1]$, with the equivalence relation $\sim$ which
        identifies $0 \sim 1$, and leaves all other points undisturbed. Then, the
        quotient space $[0, 1]/\!\sim$ is homeomorphic to the circle $S^1$.

        Note that the quotient map on $[0, 1]$ is not open, since the image of the
        open set $[0, 1 / 2)$ is not open in $[0, 1]/\!\sim$.
    \end{example}
    \begin{example}
        Let $X = \R^{n + 1}\setminus\{0\}$, and define an equivalence relation on $X$
        which identifies points on the same line through the origin together. Then,
        the resulting quotient space is called the real projective space, denoted
        $\RP^n$.
    \end{example}
    \begin{example}
        Let $S^n$ denote the $n$-sphere in $\R^{n + 1}$, and define an equivalence
        relation on $S^n$ which identifies antipodal points. Then, the resulting
        quotient space is also $\RP^n$. The quotient map here is an open map.
    \end{example}

    \begin{lemma}
        Let $f\colon X \to Y$ be an open, continuous, surjective map. Define an
        equivalence relation $\sim$ on $X$ such that $x \sim x' \Leftrightarrow
        f(x) = f(x')$. Then, $g\colon X/\!\sim \to Y$, $[x] \mapsto f(x)$ is a
        homeomorphism.
    \end{lemma}
    \begin{example}
        By defining $f$ as a composition of maps $\R^n\setminus\{0\} \to S^n \to
        S^n/\!\sim$, it can be shown that $\RP^n$ is compact.
    \end{example}


    \subsection{One-point compactification}

    \begin{definition}
        Let $X$ be a compact topological space, and let $A \subset X$ be closed. The
        one-point compactification of $X\setminus A$ is defined by \[
            Y = (X \setminus A) \cup\{\infty\},
        \] with the topology \[
            \tau_Y = \{U\subseteq X\setminus A\text{ is open}\} \cup \{Y\setminus
            C\text{ where } C\text{ is compact in }X\setminus A\}.
        \] 
    \end{definition}

    \begin{lemma}
        If $X$ is compact, Hausdorff, then so is the one-point compactification $Y$
        of $X\setminus A$.
    \end{lemma}

    \begin{lemma}
        Let $X$ be a compact, Hausdorff space and $A\subseteq X$ be a closed set.
        Define $\sim$ on $X$ by identifying $x \sim x'$ whenever $x, x' \in A$ and
        leaving the remaining points undisturbed. Then, $X/\!\sim$ is homeomorphic to
        the one-point compactification $Y = X\setminus A \cup \{\infty\}$.
    \end{lemma}
    

    \section{Countability and separation axioms}
    
    \subsection{First countability}
    
    \begin{definition}
        Let $X$ be a topological space. A countable basis at a point $x \in X$ is a
        countable collection $\beta$ of neighbourhoods of $x$ such that for any
        neighbourhood $U$ of $x$, there is a basis element $B \in \beta$ such that $x
        \in B \subseteq U$.
    \end{definition}

    \begin{definition}
        A topological space $X$ in which every element $x \in X$ admits a countable
        basis is called a first countable space.
    \end{definition}
    \begin{example}
        All metrizable spaces are first countable. Given an element $x$, the
        collection of all open balls centred at $x$ with rational radii forms a
        countable basis.
    \end{example}
    \begin{example}
        The space $\R_\ell$ is a non-metrizable space which is first countable.
    \end{example}
    
    \begin{lemma}
        The sequence lemma holds for first countable spaces, i.e.\ if $X$ is first
        countable, $A \subseteq X$, and $x \in \overline{A}$, then there exists a
        sequence $\{x_n\}_{n \in \N}$, $x_n \in A$, such that $x_n \to x$.
    \end{lemma}

    \begin{example}
        The space $\R^\omega$ with the box topology is not first countable.
    \end{example}


    \subsection{Second countability}
    
    \begin{definition}
        A topological space $X$ which admits a countable basis is called a second
        countable space.
    \end{definition}
    \begin{example}
        The Euclidean spaces $\R^n$ are second countable. The collection of all open
        balls with rational radii, centred at rational points, forms a countable basis.
    \end{example}
    \begin{example}
        The space $\R^\omega$ with the product topology is second countable.
    \end{example}
    \begin{lemma}
        If a topological space $X$ is second countable, then any discrete subspace
        $A\subseteq X$ must be countable.
    \end{lemma}

    \begin{proof}
        Let $\beta = \{B_n\}_{n \in \N}$ be a countable basis of $X$. For each $a \in
        A$, note that $\{a\}$ is open in subspace topology $A$, hence there exists a
        basis element $B_a \in \beta$ such that $B_a \cap A = \{a\}$. Furthermore,
        this assignment $A \to \beta$, $a \mapsto B_a$ is injective, hence $A$ must
        be countable.
    \end{proof}

    \begin{example}
        The space $\R^\omega$ with the uniform topology is metrizable hence first
        countable, but not second countable. This topology is induced by the metric
        \[
            \rho(x, y) = \sup_{i \in \N} \bar{d}(x_i, y_i) = \sup_{i \in \N}
            \min\{|x_i - y_i|, 1\}.
        \] Consider the subspace $A \subset \R^\omega$, consisting of all binary
        sequences. This is clearly an uncountable set. However, for any two distinct
        members $x, y \in A$, we have $d(x, y) = 1$. This precisely describes the
        discrete topology on $A$. The contrapositive of the above lemma now shows
        that $\R^\omega$ with the uniform topology cannot be second countable.
    \end{example}

    \begin{lemma}
        Let $X$ be a second countable space. Then, every open cover of $X$ admits a
        countable subcover.
        \begin{remark}
            A topological space in which every open cover admits a countable subcover
            is called a Lindel\"of space.
        \end{remark}
    \end{lemma}

    \begin{example}
        The space $\R_\ell$ is non-metrizable, first countable, and Lindel\"of, but
        not second countable. To see the latter, let $\beta$ be a basis of $\R_\ell$.
        Note that every $x \in [x, x + 1) \subset \R_\ell$, hence there must exist
        $B_x \in \beta$, $x \in B_x \subseteq [x, x + 1)$. Now, the assignment $x
        \mapsto B_x$ is injective, hence $\beta$ must be uncountable.
    \end{example}
    
    \begin{lemma}
        Let $X$ be a second countable space. Then, there exists a countable subset
        which is dense in $X$.
        \begin{remark}
            A topological space in which there exists a dense countable subset is
            called a separable space.
        \end{remark}
    \end{lemma}
    \begin{proof}
        Let $\{B_n\}_{n \in X}$ be a countable basis of $X$. Pick one element $x_n
        \in B_n$ for each $n \in \N$, whence the set $\{x_n\}_{n \in \N}$ is
        countable and dense in $X$.
    \end{proof}


    \begin{lemma}
        Subspaces, countable products of first/second countable spaces are
        first/second countable.
    \end{lemma}

    \begin{example}
        The space $\R_\ell^2$, called the Sorgenfrey plane, is not Lindel\"of, even
        though $\R_\ell$ is. This can be shown by considering the line $L = \{(x,
        -x): x \in \R\}$. Note that $L \subset \R_\ell^2$ is closed, hence $\R_\ell^2
        \setminus L$ is open. Start with this, and add the sets $[x, x + 1)\times [x,
        x + 1)$ to our collection. This is an open cover of $\R_\ell^2$ which admits
        no countable subcover.
    \end{example}


    \subsection{Separation axioms}

    \begin{definition}
        A topological space $X$ in which any two distinct points $x, y \in X$ admit
        open sets $U, V$ such that $x \in U$, $y \in V$, $U \cap V = \emptyset$, is
        called a Hausdorff space.
    \end{definition}

    \begin{definition}
        A topological space $X$ in which any point $x \in X$ and a closed set
        $F\subseteq X$ (not containing $x$) can be separated is called a regular
        space.
    \end{definition}
    
    \begin{definition}
        A topological space $X$ in which any point two disjoint closed sets $F, F'
        \subseteq X$ can be separated is called a normal space.
    \end{definition}

    \begin{lemma}
        Consider topological spaces in which singleton sets are closed. Then, all
        such normal spaces are regular, and all such regular spaces are Hausdorff.
    \end{lemma}

    \begin{example}
        Consider the space $\R_K$, and note that $K$ is closed. However, there is no
        separation of $0 \in \R$ and $K \subset \R$, hence $\R_K$ is not regular.
    \end{example}

    \begin{lemma}
        A space $X$ is regular if and only if given any point $x$ and an open set $U
        \subseteq X$ such that $x \in U$, there exists an open set $V$ such that $x
        \in V \subseteq \overline{V} \subseteq U$.
    \end{lemma}
    \begin{example}
        All metric spaces are regular.
    \end{example}

    \begin{lemma}
        Products and subspaces of regular/Hausdorff spaces are also
        regular/Hausdorff.
    \end{lemma}
    \begin{example}
        The space $\R_\ell$ is regular, but not metrizable, and so is $\R_\ell^2$.
    \end{example}

    \begin{lemma}
        All metrizable spaces are normal.
    \end{lemma}
    \begin{proof}
        Let $X$ be metrizable, and let $A, B \subseteq X$ be closed, $A \cap B =
        \emptyset$. Then for each $a \in A$, $b \in B$, we can choose $\epsilon_a,
        \epsilon_b > 0$ such that \[
            B(a, \epsilon_a) \cap B = \emptyset, \qquad
            B(b, \epsilon_b) \cap A = \emptyset.
        \] This is because the complements $X\setminus A$, $X\setminus B$ are open.
        Now set \[
            U = \bigcup_{a \in A} B(a, \epsilon_a / 2), \qquad
            V = \bigcup_{b \in A} B(b, \epsilon_b / 2).
        \] It can be checked that these are open, with $A \subseteq U$, $B \subseteq
        V$, $U \cap V = \emptyset$.
    \end{proof}
    
    \begin{example}
        The space $\R_\ell$ is normal, but not metrizable. However, $\R_\ell^2$ is
        not normal.
    \end{example}

    \begin{theorem}
        All compact, Hausdorff spaces are normal.
    \end{theorem}

    \begin{theorem}
        All regular, second countable spaces are normal.
    \end{theorem}

    \begin{example}
        The space $\R_\ell$ is normal but not second countable.
    \end{example}

    \begin{example}
        The Sorgenfrey line $L \subset \R_\ell^2$ has the discrete topology. Then,
        any subset $A \subseteq L$ is both open and closed, and $A, L\setminus A$ are
        closed in $\R_\ell^2$. If $\R_\ell^2$ were normal, then we could separate
        them in $\R_\ell^2$ by open sets $U_A, V_A$ with $A \subseteq U_A$,
        $L\setminus A \subseteq V_A$, $U_A \cap V_A = \emptyset$. Let $D \subset
        \R_\ell^2$ be the set of all points with rational coordinates; this is dense
        in $\R_\ell^2$. Define \[
            \theta\colon \mathcal{P}(L) \to \mathcal{P}(D), \qquad 
            A \mapsto \begin{cases}
                D \cap U_A, &\text{ if } \emptyset \subset A \subset L, \\
                D, &\text{ if } A = L, \\
                \emptyset, &\text{ if } A = \emptyset.
            \end{cases}
        \] Note that $U_A \cap D$ is non-empty when $A \neq \emptyset$ by the density
        of $D$. Pick $A \neq B$, then there exists $x \in A, x \notin B$ hence $x \in
        L \setminus B$ or $x \in U_A \cap V_B$. Thus, $D \cap U_A$ and $D \cap V_B$
        are non-empty. There must be points in $D \cap U_A$ not in $D \cap U_B$,
        hence $\theta$ is injective. However, this is already a contradiction, via
        cardinalities.
    \end{example}

    \begin{example}
        Let $X$ be a metric space, and let $A, B \subseteq X$ be closed and disjoint.
        We can define \[
            f\colon X \to \R, \qquad x \mapsto \frac{d(x, A)}{d(x, A) + d(x, B)}.
        \] Then, $f$ is continuous, with $f(A) = 0$ and $f(B) = 1$. This is a
        `functional separation' of $A$ and $B$.

        Note that if $A$ and $B$ were not closed, the denominator might have been
        zero.
    \end{example}


    \begin{lemma}[Urysohn]
        A space $X$ is normal if and only if given any two non-empty closed disjoint
        sets $A, B \subseteq X$, there exists a continuous function $f\colon X \to
        [0, 1]$ such that $f(A) = 0$, $f(B) = 1$.
    \end{lemma}

    \begin{definition}
        Let $X, Y$ be topological spaces. An imbedding $f\colon X \to Y$ is a
        homeomorphism between $X$ and $f(X)$.
    \end{definition}

    \begin{theorem}[Urysohn metrization theorem]
        All regular, second countable spaces are metrizable.
    \end{theorem}



    \subsection{Compact manifolds}
    
    \begin{definition}
        An $n$-manifold is a regular, second-countable space such that each point has
        a neighbourhood homeomorphic to an open subset of $\R^n$.
    \end{definition}
    \begin{example}
        The real projective space $\RP^n$ is a compact manifold. Furthermore, it can
        be covered by finitely many open sets, each of which is homeomorphic to an
        open subset of $\R^n$. To see this, define \[
            U_i = \{[x] : x \in \R^{n + 1},  x_i \neq 0\} \subset \RP^n.
        \] It is clear that the collection of $U_i$ cover $\RP^n$. Now, define the
        maps \[
            \varphi_i\colon U_i \to \R^n, \qquad [x] \mapsto \left(\frac{x_1}{x_i},
            \dots, \frac{x_{i - 1}}{x_i}, \frac{x_{i + 1}}{x_i}, \dots,
            \frac{x_n}{x_i}\right).
        \] Note that $\varphi_i$ is injective; if $\varphi_i([x]) = \varphi_i([y])$,
        then each $y_j/y_i = x_j/x_i$, or $y_j = \lambda x_j$, but this is precisely
        what it means for $x \sim y$, hence $[x] = [y]$. The map $\varphi_i$ is also
        continuous, since the maps \[
            x \mapsto \left(\frac{x_1}{x_i}, \dots, \frac{x_{i - 1}}{x_i}, \frac{x_{i
            + 1}}{x_i}, \dots, \frac{x_n}{x_i}\right).
        \] are continuous, and the quotient map $\R^{n + 1}\setminus\{0\} \to \RP^n$
        is open. It can also be shown that $\varphi_i^{-1}$ is continuous, hence the
        maps $\varphi_i$ are homeomorphisms.
    \end{example}

    \begin{definition}
        Let $X$ be a topological space, and let $\mathscr{U} = \{U_1, \dots, U_k\}$
        be an open cover. A partition of unity, dominated by $\mathscr{U}$, is a
        collection of continuous real functions $\psi_1, \dots, \psi_k$ on $X$
        such that each $\supp{\psi_i} \subseteq U_i$ and $\sum_{i = 1}^k \psi_i =
        1$.
    \end{definition}

    \begin{lemma}
        Let $X$ be a topological space, and let $\mathscr{U} = \{U_1, \dots, U_k\}$
        be an open cover. If $X$ is normal, then there exists a partition of unity
        dominated by $\mathscr{U}$.
    \end{lemma}
    \begin{proof}
        First, we refine the open cover $\mathscr{U}$ and extract an open cover
        $\mathscr{V} = \{V_1, \dots, V_k\}$ such that each $V_i \subseteq
        \overline{V_i} \subseteq U_i$. To do so, first define \[
            A_1 = X \setminus (U_2 \cup \dots \cup U_k).
        \] Then, $A_1 \subseteq U_1$ is closed in $X$. By the normality of $X$, there
        exists open $V_1$ such that $A_1 \subseteq V_1 \subseteq \overline{V_1}
        \subseteq U_1$. Now, the collection $\{V_1, U_2, \dots, U_k\}$ is still an
        open cover of $X$. This process can be repeated; if at any point we have an
        open cover $\{V_1, \dots, V_j, U_{j + 1}, \dots, U_k\}$, set \[
            A_{j + 1} = X\setminus (V_1 \cup \dots \cup V_j \cup U_{j + 2} \cup \dots
            \cup U_k),
        \] and argue as before to obtain open $V_{j + 1} \subseteq A_{j + 1}$, and
        the open cover $\{V_1, \dots, V_{j + 1}, U_{j + 1}, \dots, U_k\}$.
        
        Next, perform another refinement of $\mathscr{V}$ and extract an open cover
        $\mathscr{W} = \{W_1, \dots, W_k\}$.

        Now, apply the Urysohn Lemma and exhibit continuous functions $\varphi_1,
        \dots, \varphi_k$ on $X$ such that each $\varphi_i(\overline{W_i}) = 1$,
        $\varphi_i(X\setminus V_i) = 0$. Note that each $\supp{\varphi_i} \subseteq
        \overline{V_i} \subseteq U_i$. Finally, set \[
            \psi_i = \frac{\varphi_i}{\Phi}, \qquad \Phi = \sum_{i = 1}^k \varphi_i.
            \qedhere
        \] 
    \end{proof}


    \begin{theorem}
        Any compact Hausdorff manifold can be imbedded into some Euclidean space.
    \end{theorem}
    \begin{proof}
        Let $\mathscr{U} = \{U_1, \dots, U_k\}$ be an open cover of $X$, and let
        $\varphi_i\colon U_i \to \varphi_i(U_i) \subseteq \R^n$ be homeomorphisms.
        Since $X$ is compact and Hausdorff, hence normal, choose a partition of unity
        $\psi_1, \dots, \psi_k$ dominated by $\mathscr{U}$. Set $A_i = \supp{\psi_i}
        \subseteq U_i$, and define \[
            h_i\colon X \to \R, \qquad x\mapsto \varphi_i(x)\,\psi_i(x).
        \] It is clear that each $h_i$ is continuous, with $h_i = 0$ on $X \setminus
        A_i$. Now, define \[
            f\colon X \to \underbrace{\R \times \dots \times \R}_{k \text{ times}}\times 
            \underbrace{\R^n \times \dots \times \R^n}_{k \text{ times}}, \qquad
            x \mapsto (\psi_1(x), \dots, \psi_k(x), h_1(x), \dots, h_k(x)).
        \] Then, $f$ is continuous as well as injective; suppose that $f(x) = f(y)$.
        Since $\mathscr{U}$ covers $X$, we must have some $\psi_j(x) > 0$; but
        $\psi_j(x) = \psi_j(y) > 0$, so $x, y \in A_j \subseteq U_j$. Thus, from
        $h_j(x) = h_j(y)$, we can cancel $\psi_j(x), \psi_j(y)$ to obtain
        $\varphi_j(x) = \varphi_j(y)$. The bijectivity of $\varphi_j$ immediately
        gives $x = y$. Thus, $f$ is a continuous map from a compact space to a
        Hausdorff space, hence $f$ is a homeomorphism onto its image. In other words,
        $f$ is an imbedding of $X$ in $\R^{(n + 1)k}$.
    \end{proof}

    \begin{theorem}
        Any compact Hausdorff space without any isolated points is uncountable.
    \end{theorem}
    \begin{proof}
        First, we show that given any $x \in X$ and an open set $U \subseteq X$, we
        can always find non-empty open $V \subseteq U$ with $x \notin V$. To show
        this, note that since $x \in U$ is not an isolated point, we must have some
        distinct point $y \in U$. Since $X$ is Hausdorff, find disjoint open sets
        $W_1, W_2$ with $x \in W_1$, $y \in W_2$. Now we can set $V = W_2 \cap U$.
        Note that we also have $x \notin \overline{V}$ since $x \in W_1 \subseteq
        W_2^c$.

        Now, we show that there is no surjection $f\colon \N \to X$. If there was
        such an $f$, denote $x_n = f(n)$. Set $U = X$, and use the above process to
        find non-empty open $V_1 \subseteq U$, $x_1 \notin \overline{V_1}$. Next, set
        $U = V_1$, and find non-empty open $V_2 \subseteq V_1$, $x_2 \notin
        \overline{V_2}$. In this manner, we can obtain a chain of non-empty closed
        sets \[
            \overline{V_1} \supseteq \overline{V_2} \supseteq \dots \supseteq
            \overline{V_i} \supseteq \dots.
        \] These clearly have the finite intersection property, but $\bigcap_{i =
        1}^\infty \overline{V_i} = \emptyset$ since each $x_i \notin V_i$. This
        contradicts the compactness of $X$.
    \end{proof}


    \begin{theorem}[Lebesgue number]
        Let $X$ be a compact metric space, and let $\mathscr{U}$ be an open cover of
        $X$. Then, there exists a corresponding positive number $\delta$ such that
        any subset of $X$ with diameter less than $\delta$ is contained in one of the
        elements of $\mathscr{U}$.
    \end{theorem}
    \begin{proof}
        Note that for any $A \subset X$, the map $x \mapsto d(x, A)$ is continuous.
        Now, using the compactness of $X$, extract a finite subcover $U_1, \dots,
        U_n$ from $\mathscr{U}$. Set $F_i = X\setminus U_i$, and define \[
            f\colon X \to \R, \qquad x \mapsto \frac{1}{n}\sum_{i = 1}^n d(x, F_i).
        \] This map is continuous, and strictly positive. Since $X$ is compact, this
        map must attain a minimum $\delta > 0$. We now show that this $\delta$ is
        indeed the Lebesgue number for the open cover $\mathscr{U}$.

        Let $B \subseteq X$, $\diam{B} < \delta$. Pick $x_0 \in B$, whence $f(x_0)
        \geq \delta$. Suppose that all $d(x_0, F_i) \leq d(x_0, F_k)$ for some $k$,
        then $\delta \leq f(x_0) \leq d(x_0, F_k)$. We have \[
            d(x_0, F_k) \geq \delta \;\implies\; \inf\{d(x_0, c) : c \notin U_k\} \geq
            \delta.
        \] If $B \not\subseteq U_k$, then we could have chosen $x \in B$, $x \notin
        U_k$. From $x \in B$, we have $d(x_0, x) < \delta$; this contradicts the
        above.
    \end{proof}
    \begin{corollary}
        Let $X$ be a compact metric space. Then, all continuous functions $f\colon X
        \to Y$, where $Y$ is a metric space, are uniformly continuous.
    \end{corollary}
    \begin{proof}
        Let $f\colon X \to Y$ be continuous, and let $\epsilon > 0$. Set $\mathscr{U}
        = \{f^{-1}(B_{\epsilon / 2}(y)) : y \in Y\}$, which is clearly an open cover
        of $X$. Pick the corresponding Lebesgue number $\delta > 0$. Now, given any
        two points $x, x' \in X$ such that $d_X(x, x') < \delta$, we have $\diam\{x,
        x'\} < \delta$ hence $\{x, x'\} \subseteq f^{-1}(B_{\epsilon / 2}(y))$ for
        some $y \in Y$.  Thus, $f(x), f(x') \in B_{\epsilon / 2}(y)$, hence
        $d_Y(f(x), f(x')) < \epsilon$.
    \end{proof}


    \section{Topological groups}

    \begin{definition}
        A topological group $G$ is a topological space that is also a group, such
        that the product and inversion maps are continuous, i.e.\ the maps $(x,
        y) \mapsto xy$ and $x \mapsto x^{-1}$ are continuous.
        \begin{remark}
            We also demand that $G$ be Hausdorff.
        \end{remark}
    \end{definition}
    \begin{example}
        The group $GL_n(\R)$ is a topological group. Here, we topologize $GL_n(\R)$
        using the Euclidean topology on $\R^{n^2}$. 
        \begin{remark}
            Note that the determinant map $\det\colon M_n(\R) \to \R$ is continuous,
            being a polynomial in the entries of any given matrix. Thus, $GL_n(\R)$
            is open in $M_n(\R)$, being the pre-image of the open set
            $\R\setminus\{0\}$.
        \end{remark}
        \begin{remark}
            Note that $GL_n(\R)$ is not connected, being the disjoint union of
            $\det^{-1}(-\infty, 0)$ and $\det^{-1}(0, \infty)$. However, $GL_n(\C)$
            is path connected. To show this, pick $A \in GL_n(\C)$, and let $z \in
            \C$ such that $|z| = 1$, with no $\lambda z$ being an eigenvalue of $A$
            (this can be done since $A$ has only finitely many eigenvalues). Now,
            consider the map \[
                \gamma\colon [0, 1] \to GL_n(\C), \qquad t \mapsto tA + (1 - t)zI_n.
            \] Note that $\gamma$ is continuous and well-defined; $\gamma(0) = I_n$,
            $\gamma(1) = A$ both of which are in $GL_n(\C)$. Now if some
            $\det\gamma(t) = 0$ for $0 < t < 1$, there must be some non-zero vector
            $v\in \C^n$ such that $tAv + (1 - t)zI_nv = 0$, i.e.\ $Av = -(1 -
            t)z/t\,v$. However, $z$ was chosen such that none of its multiples are
            eigenvectors of $A$. Thus, $\gamma$ is a path joining $I_n$ with $A$,
            which shows that $GL_n(\C)$ is indeed path connected.
        \end{remark}
    \end{example}
    \begin{example}
        The topological group of unitary matrices $U_n(\C)$ is compact. Note that it
        is sufficient to show that $U_n(\C)$ is closed and bounded. First, the maps
        $A \mapsto A A^* - I_n$ and $A \mapsto A^* A - I_n$ are both continuous,
        hence $U_n(\C)$ is closed, being the intersection of the pre-images of
        $\{0\}$ under these maps. Next, note that given $U \in U_n(\C)$, we have $U
        U^* = I_n$, hence the columns of $U$ are orthonormal. Thus, \[
            \sum_{i, j} |u_{ij}|^2 = n,
        \] showing that $U$ is bounded.
    \end{example}

    \begin{theorem}
        Let $G$ be a topological group, and let $H$ be a closed normal subgroup of
        $G$. If $G/H$ and $H$ are connected, then $G$ is connected.
    \end{theorem}
    \begin{proof}
        Suppose that $G$ is disconnected. Then there exists a continuous, surjective
        map $\Phi\colon G \to \{0, 1\}$. Since $H \subset G$ is connected, one of
        $\Phi(H) = 0, 1$. Now, the quotient map $\pi\colon G \to G/H$ induces the
        continuous surjection $\tilde{\Phi}\colon G/H \to \{0, 1\}$, $\Phi =
        \tilde{\Phi} \circ q$, which contradicts the connectedness of $G/H$. Note
        that we have used the fact that $\pi$ is open; to see this, if $U\subseteq G$
        is open, we claim that $\pi(U) \subseteq G/H$ is open, i.e.\
        $\pi^{-1}(\pi(U))$ is open in $G$. But, \[
            \pi(U) = \{gH : g \in U\},
        \] so \begin{align*}
            \pi^{-1}(\pi(U)) \;&=\; \{g \in G: gH = g'H, g' \in U\} \\
            \;&=\; \{g \in G : g'^{-1}g \in H, g' \in U\} \\
            \;&=\; \bigcup_{h \in H} Uh.
        \end{align*}
        which is open in $G$.
    \end{proof}

\end{document}
