---
title: "Assignment 2b"
author: "Satvik Saha"
date: "`r Sys.Date()`"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Answer 1

```{r 1}
v1 <- rep(1, 10)
v2 <- 1:10
v3 <- v2^2

X  <- rbind(v1, v2, v3)
```

(a) For equal weights, the weighted average is just the usual simple average.
```{r 1a}
colMeans(X)
```

(b)
```{r 1b}
w <- c(1, 2, 3)
colSums(X * w) / sum(w)
```


## Answer 2

(a) Number of games won by a chess player who has played 20 games, with a probability of 30% of winning each game.
```{r 2a}
rbinom(1, 20, 0.3)
```

(b) Number of games won by a couple of chess players who have played 20 games (not against each other). The first has a probability of 30% of winning each game, while the second has a probability of 40% of winning each game.
```{r 2b}
rbinom(2, 20, c(0.3, 0.4))
```

(c) Number of games won by a couple of chess players. The first has played 30 games with a probability of 30% of winning each game, while the second has played 20 games with a probability of 40% of winning each game.
```{r 2c}
rbinom(2, c(30, 20), c(0.3, 0.4))
```