---
title: "Assignment 9a"
author: "Satvik Saha"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.pos = "!H", out.extra = "")
```

## Answer 1

(a) The new standard error with a sample size of $n$ will be around $se = 4.0
    \sqrt{100 / n}$.
    We want $2.8se = 5.0 - 1.4$, hence $n \approx (2.8\cdot 4.0 / (5.0
    - 1.4))^2 \cdot 100 \approx `r ceiling((2.8 * 4.0 / (5.0 - 1.4))^2
    * 100)`$.

(b) This time, we want $5.0 - 1.4$ to be
    ``` {r 1b}
    qnorm(0.8) + qnorm(0.95)
    ```
    standard errors, so $n \approx (2.5\cdot 4.0 / (5.0 - 1.4))^2 \cdot 100
    \approx `r ceiling((2.5 * 4.0 / (5.0 - 1.4))^2 * 100)`$.



## Answer 2

(a) We equate $0.03 = 0.5 / \sqrt{n}$, whence $n \approx (0.5 / 0.03)^2 \approx
    `r ceiling((0.5 / 0.03)^2)`$.

(b) We make the approximation that $14\%$ of our sample will be Latinos, hence
    we ask for $n \approx (0.5 / 0.03)^2 / 0.14 \approx `r ceiling((0.5
    / 0.03)^2 / 0.14)`$.

(c) We approximate $0.01 = \sqrt{0.14\cdot (1 - 0.14) / n}$, whence $n \approx
    0.14 \cdot (1 - 0.14) / 0.01^2 \approx `r ceiling(0.14 * (1 - 0.14) / 0.01^2)`$.




## Research homework assignment

Suppose that we are in the setup $y_i \sim N(g(x_i), \sigma^2)$, and $x_i$ take
values in $0, 0.5, 1$ with frequencies $(1 - w)n / 2, wn, (1 - w)n / 2$.
Then, for $x \in \{0, 0.5, 1\}$, we have the means within each group $\bar{y}_x
\sim N(g(x), \sigma^2 / n_x)$.
In order to estimate the full effect $\theta_i = g(1) - g(0)$ and the
half-effect $\theta_{0.5} = g(0.5) - g(0)$, suppose that we use $\hat{\theta}_i
= \bar{y}_1 - \bar{y}_0$ and $\hat{\theta}_{0.5} = \bar{y}_{0.5} - \bar{y}_0$.
We are also interested in the relative non-linearity $\delta = (0.5\theta_1
- \theta_{0.5}) / \theta_1$.

Check that $$
    \text{MSE}(\hat{\theta}_1) = \frac{4}{1 - w} \frac{\sigma^2}{n}, \qquad
    \text{MSE}(\hat{\theta}_{0.5}) = \frac{1 + w}{w (1 - w)} \frac{\sigma^2}{n}.
$$
When $w = 0$, we have $$
    \text{MSE}(\hat{\theta}_1^0) = 4 \,\frac{\sigma^2}{n}.
$$
When $w = 1 / 3$, we have $$
    \text{MSE}(\hat{\theta}_1^{1/3}) = 6 \,\frac{\sigma^2}{n}, \qquad
    \text{MSE}(\hat{\theta}_{0.5}^{1/3}) = 6 \,\frac{\sigma^2}{n}.
$$

We would like to measure the deviation from linearity $\Delta = 0.5\theta_1
- \theta_{0.5}$, estimated by $\hat{\Delta} = 0.5\hat{\theta}_1
- \hat{\theta}_{0.5}$, which has $$
    \text{MSE}(\hat{\Delta}^{1/3}) = 7.5 \,\frac{\sigma^2}{n}.
$$

Thus, if we are satisfied with the estimate of the full effect $\theta_1$ with
the scheme $w = 0$ for some sample size $n$, say with our MSE $4\sigma^2
/ n < \epsilon$, then we would need a sample size of around $2n$ with the
scheme $w = 1/3$ for our MSE for the estimated deviation from linearity
$\Delta$ to be similarly small, since $7.5 \sigma^2 / 2n < \epsilon$.
