---
title: "Assignment 9b"
author: "Satvik Saha"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.pos = "!H", out.extra = "")
```

## Answer 1

``` {r 1 data}
df <- data.frame(
    i = c("Audrey", "Anna", "Bob", "Bill", "Caitlin", "Cara", "Dave", "Doug"),
    x = c(40, 45, 50, 55, 60, 65, 70, 75),
    z = c(0, 1, 1, 0, 0, 0, 1, 1),
    y0 = c(140, 140, 150, 150, 160, 160, 170, 170),
    y1 = c(135, 135, 140, 140, 155, 155, 160, 160)
)
```
(a)
    ``` {r 1a}
    audrey <- df$i == "Audrey"
    df$y1[audrey] - df$y0[audrey]
    ```
    Thus, the treatment effect for Audrey is $-5$.

(b)
    ``` {r 1b}
    df$tau <- df$y1 - df$y0
    mean(df$tau)
    ```
    Thus, the sample average treatment effect is $-7.5$.

(c)
    ``` {r 1c}
    df$y <- ifelse(df$z == 0, df$y0, df$y1)
    lm(y ~ z, data = df)
    ```
    Thus, the estimated treatement effect is $-3.75$.


    For the sake of curiosity, consider
    ``` {r 1c remark}
    lm(y ~ z + x, data = df)
    ```



## Answer 2

It would seem that (b) is true; the prediction $X\beta$ has a smoothing effect
leading to the absence of more extreme values for income.
In particular, the imputed data will be 'missing' extreme values produced by
the error terms.
